---
title: "Alzheimer's Disease Prediction Using DNA Methylation Markers"
author: "Marco Banfi, Simone Carrieri, Tomás Serra"
date: today
title-slide-attributes:
  data-background-color: "#F2C94C"
format:
  revealjs:
    slide-number: true
    chalkboard: true
    preview-links: auto
    footer: "Bioinformatics Project - AD Prediction"
    transition: slide
    background-transition: fade
    css: styles.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, 
                      fig.width = 10, fig.height = 6)
```

# 1. Introduction 

---

## Project Overview {.smaller}

**Goal:** Develop a predictive model for Alzheimer's Disease using DNA methylation data

::: {.incremental}
- This project develops a predictive model for Alzheimer's Disease (AD) using DNA methylation data from post-mortem brain tissue.
- After controlling for demographic and technical confounders we use logistic regression to identify which methylation markers can distinguish AD patients from healthy controls.
:::

---

## Alzheimer's Disease Overview {.smaller}

:::: {.columns}
::: {.column width="60%"}
**Clinical Features:**

- Progressive neurodegenerative disorder
- Memory loss and cognitive decline
- Non-monogenic (complex genetics)

**Pathological Hallmarks:**

- Amyloid-beta plaques
- Neurofibrillary tangles
- Neuronal loss
:::

::: {.column width="40%"}
**Braak Staging (0-6):**

- **0:** No pathology
- **I-II:** Transentorhinal
- **III-IV:** Limbic
- **V-VI:** Neocortical (severe)
:::
::::

---

## DNA Methylation Basics {.smaller}

**What is DNA methylation?**

Epigenetic modification where methyl groups (CH₃) are added to cytosine bases at CpG dinucleotides

**Why it matters in AD:**

::: {.incremental}
- **Epigenetic dysregulation:** AD brains show altered methylation patterns
- **Gene expression changes:** Affects inflammation, synaptic function, neuronal survival
- **Potential biomarkers:** Disease indicators and therapeutic targets
:::

---

## Key AD-Related Genes {.smaller}

Recent studies identified differential methylation in:

:::: {.columns}
::: {.column width="50%"}
- **APOE** (apolipoprotein E)
  - Major genetic risk factor
  
- **APP** (amyloid precursor protein)
  - Involved in plaque formation
:::

::: {.column width="50%"}
- **MAPT** (microtubule-associated protein tau)
  - Forms tangles
  
- **Inflammatory response genes**
  - Multiple pathways affected
:::
::::

---

# 2. Dataset & Preprocessing

---

## Dataset: GSE66351 {.smaller}

**Source:** Gene Expression Omnibus (GEO)

**Platform:** Illumina HumanMethylation450 BeadChip

**Citation:** Lunnon et al. (2014) - "Methylomic profiling implicates cortical deregulation of ANK1 in Alzheimer's disease"

**Sample Size:** 190 post-mortem brain tissue samples

---

## Clinical Variables {.smaller}

```{r install-packages, results='hide', message=FALSE}
# Install required packages
if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
if (!require("GEOquery", quietly = TRUE))
  BiocManager::install("GEOquery")
if (!require("caret", quietly = TRUE))
  install.packages("caret")

library(GEOquery)
library(dplyr)
library(ggplot2)
library(caret)
set.seed(123)
```

```{r download-data, cache=TRUE, message=FALSE, warning=FALSE}
options(timeout = 6000)
data_dir <- "data"
if (!dir.exists(data_dir)) dir.create(data_dir)
gse_file <- file.path(data_dir, "GSE66351.rds")

if (!file.exists(gse_file)) {
  gse <- GEOquery::getGEO("GSE66351", GSEMatrix = TRUE, 
                          AnnotGPL = FALSE, getGPL = FALSE)[[1]]
  saveRDS(gse, gse_file)
} else {
  gse <- readRDS(gse_file)
}

pheno_data <- pData(phenoData(gse))
methylation_data <- exprs(gse)
```

1. **Diagnosis:** AD vs Control (outcome variable)
2. **Age:** Patient age at death (continuous)
3. **Sex:** Male or Female (categorical)
4. **Braak stage:** AD progression severity (0-6)
5. **Cell type:** Neuronal, glial, or bulk tissue
6. **Brain region:** Cortical region sampled

---

## Sample Characteristics {.smaller}

```{r clinical-data}
clinical_data <- data.frame(
  sample_id = pheno_data$geo_accession,
  diagnosis = gsub("disease state: ", "", gsub(".*: ", "", pheno_data$characteristics_ch1.1)),
  age = as.numeric(gsub(".*: ", "", pheno_data$characteristics_ch1.4)),
  sex = gsub(".*: ", "", pheno_data$characteristics_ch1.5),
  cell_type = gsub(".*: ", "", pheno_data$characteristics_ch1),
  braak_stage = gsub(".*: ", "", pheno_data$characteristics_ch1.2),
  region = gsub(".*: ", "", pheno_data$characteristics_ch1.3)
)

clinical_data$AD_status <- ifelse(clinical_data$diagnosis == "AD", 1, 0)
clinical_data$sex <- factor(clinical_data$sex)
clinical_data$cell_type <- factor(clinical_data$cell_type)
clinical_data$region <- factor(clinical_data$region)

summary_stats <- list(
  n_total = nrow(clinical_data),
  n_ad = sum(clinical_data$AD_status),
  n_control = sum(clinical_data$AD_status == 0),
  age_mean = round(mean(clinical_data$age, na.rm=TRUE), 1),
  age_range = range(clinical_data$age, na.rm=TRUE)
)
```

**Clinical Summary:**

- Total samples: `r summary_stats$n_total`
- AD cases: `r summary_stats$n_ad`
- Controls: `r summary_stats$n_control`
- Mean age: `r summary_stats$age_mean` years (range: `r summary_stats$age_range[1]`-`r summary_stats$age_range[2]`)

---

## Distribution Tables {.smaller}

:::: {.columns}
::: {.column width="50%"}
```{r}
knitr::kable(table(clinical_data$sex), col.names = c("Sex", "N"))
```
:::

::: {.column width="50%"}
```{r}
knitr::kable(table(clinical_data$cell_type), col.names = c("Cell Type", "N"))
```
:::
::::

```{r}
knitr::kable(table(clinical_data$braak_stage), col.names = c("Braak Stage", "N"))
```

---

## Beta Values vs M-Values {.smaller}

**The Problem with Beta Values (β ∈ [0,1]):**

::: {.incremental}
- **Heteroscedasticity:** Variance not constant across range
- **Boundary constraints:** Can't use standard linear models effectively
- **Non-normal distribution:** Often bimodal
:::

. . .

**The M-Value Solution:**

$$M = \log_2\left(\frac{\beta}{1-\beta}\right)$$

::: {.incremental}
- Unbounded range (-∞ to +∞)
- Homoscedastic (constant variance)
- Better for linear modeling
:::

---

## M-Value Interpretation {.smaller}

```{r beta-to-m}
beta_to_m <- function(beta) {
  beta[beta <= 0] <- 0.001
  beta[beta >= 1] <- 0.999
  m_value <- log2(beta / (1 - beta))
  return(m_value)
}
m_values <- beta_to_m(methylation_data)
```

**Interpretation:**

- M = 0: Equal methylated/unmethylated (β = 0.5)
- M > 0: More methylated than unmethylated
- M < 0: More unmethylated than methylated
- Difference of 1 M-value ≈ 2-fold change in methylation ratio

**Our data:**

- Range: `r paste(round(range(m_values, na.rm=TRUE), 2), collapse=" to ")`
- Median: `r round(median(m_values, na.rm=TRUE), 2)`

---

## Beta vs M-Value Distributions {.smaller}

```{r plot-beta-vs-m, fig.height=5}
set.seed(123)
sample_idx <- sample(length(methylation_data), 10000)
sample_beta <- as.vector(methylation_data)[sample_idx]
sample_m <- as.vector(m_values)[sample_idx]

par(mfrow=c(1,2))
hist(sample_beta, breaks=50, main="Beta Values", xlab="Beta", col="lightblue")
hist(sample_m, breaks=50, main="M-Values", xlab="M-value", col="orange")
par(mfrow=c(1,1))
```

---

# 3. Feature Selection

---


## The Dimensionality Problem {.smaller}

**Challenge:** 481,868 CpG sites but only 190 samples

**Problems this creates:**

::: {.incremental}
1. **Overfitting:** Model memorizes instead of learning
2. **Computational burden:** Testing all sites impractical
3. **Multiple testing:** Hundreds of thousands of tests
4. **Loss of power:** Irrelevant features obscure true signals
:::

---

## Solution: Two-Stage Analysis {.smaller}

**Stage 1: Feature Selection**

- Use linear regression to find CpG sites associated with AD progression
- Apply stringent multiple testing correction (Bonferroni)

. . .

**Stage 2: Predictive Modeling**

- Use selected CpGs in logistic regression
- Predict AD diagnosis (binary outcome)
- Validate on test set

---

## Linear Model for Feature Selection {.smaller}

For each CpG site:

$$M_{ij} = \beta_0 + \beta_1 \cdot Braak_i + \beta_2 \cdot Age_i + \beta_3 \cdot Sex_i + \beta_4 \cdot CellType_i + \beta_5 \cdot Region_i + \epsilon_{ij}$$

**Why this approach?**

::: {.incremental}
1. Tests association with disease **progression** (more sensitive)
2. Standard in EWAS (Epigenome-Wide Association Studies)
3. Identifies mechanistic changes, not just group differences
:::

---

## Why Include Covariates? {.smaller}

**Without covariates:** Too many false positives due to confounding

:::: {.columns}
::: {.column width="50%"}
**Potential Confounders:**

- Age differences
- Sex differences
- Cell type composition
- Brain region effects
:::

::: {.column width="50%"}
**With covariates:**

✓ True AD-specific effects

✓ Mechanistic insights

✓ Better generalization

✓ Reduced false positives
:::
::::

---

## Multiple Testing Correction {.smaller}

**The Problem:** Testing 481,868 sites at α = 0.05

Expected false positives: 481,868 × 0.05 = **24,093 false positives!**

. . .

**Two Approaches:**

1. **False Discovery Rate (FDR):**
   - Controls proportion of false discoveries
   - FDR < 0.05 means < 5% of significant findings are false positives
   - Less conservative, higher power

2. **Bonferroni Correction:**
   - Controls family-wise error rate
   - Adjusted p-value: 0.05 / 485,577 = 1.03 × 10⁻⁷
   - More conservative, stricter control

---

## Fitting Linear Models {.smaller}

```{r feature-selection-setup}
braak_numeric <- as.numeric(clinical_data$braak_stage)
valid_idx <- !is.na(braak_numeric)
clinical_data_valid <- clinical_data[valid_idx, ]
m_values_valid <- m_values[, valid_idx]
```

```{r fit-linear-models, cache=TRUE, results='hide', message=FALSE}
fit_cpg_model_covariates <- function(m_vals, clinical) {
  braak <- as.numeric(clinical$braak_stage)
  df <- data.frame(
    m_value = m_vals, braak_stage = braak, age = clinical$age,
    sex = clinical$sex, cell_type = clinical$cell_type, region = clinical$region
  )
  df <- df[complete.cases(df), ]
  if (nrow(df) < 20 || length(unique(df$braak_stage)) < 3) {
    return(list(beta = NA, se = NA, t_stat = NA, p_value = NA, 
                adj_r_squared = NA, n_samples = nrow(df)))
  }
  tryCatch({
    model <- lm(m_value ~ braak_stage + age + sex + cell_type + region, data = df)
    coef_summary <- summary(model)$coefficients
    if (!"braak_stage" %in% rownames(coef_summary)) {
      return(list(beta = NA, se = NA, t_stat = NA, p_value = NA, 
                  adj_r_squared = NA, n_samples = nrow(df)))
    }
    braak_row <- which(rownames(coef_summary) == "braak_stage")
    return(list(
      beta = coef_summary[braak_row, "Estimate"],
      se = coef_summary[braak_row, "Std. Error"],
      t_stat = coef_summary[braak_row, "t value"],
      p_value = coef_summary[braak_row, "Pr(>|t|)"],
      adj_r_squared = summary(model)$adj.r.squared,
      n_samples = nrow(df)
    ))
  }, error = function(e) {
    return(list(beta = NA, se = NA, t_stat = NA, p_value = NA, 
                adj_r_squared = NA, n_samples = nrow(df)))
  })
}

cache_file <- file.path(data_dir, "cpg_results.rds")
if (file.exists(cache_file)) {
  cpg_results <- readRDS(cache_file)
} else {
  results_list <- list()
  for (i in 1:nrow(m_values_valid)) {
    results_list[[i]] <- fit_cpg_model_covariates(m_values_valid[i, ], clinical_data_valid)
  }
  cpg_results <- data.frame(
    CpG = rownames(m_values_valid),
    Beta_Coefficient = sapply(results_list, function(x) x$beta),
    Std_Error = sapply(results_list, function(x) x$se),
    t_statistic = sapply(results_list, function(x) x$t_stat),
    p_value = sapply(results_list, function(x) x$p_value),
    adj_r_squared = sapply(results_list, function(x) x$adj_r_squared),
    n_samples = sapply(results_list, function(x) x$n_samples)
  )
  cpg_results <- cpg_results[complete.cases(cpg_results), ]
  saveRDS(cpg_results, cache_file)
}
```

**Progress:**

- Successfully analyzed: `r nrow(cpg_results)` CpG sites
- Failed (insufficient data): `r nrow(m_values_valid) - nrow(cpg_results)`
- Beta_Coefficient = M-value change per Braak stage increase
- Controlled for age, sex, cell type, and region

---

## Correction Results {.smaller}

```{r multiple-testing}
cpg_results$p_bonferroni <- p.adjust(cpg_results$p_value, method = "bonferroni")
cpg_results$p_fdr <- p.adjust(cpg_results$p_value, method = "fdr")

n_sig_raw <- sum(cpg_results$p_value < 0.05)
n_sig_fdr <- sum(cpg_results$p_fdr < 0.05)
n_sig_bonf <- sum(cpg_results$p_bonferroni < 0.05)
```

**Significance Testing Results:**

- Uncorrected p < 0.05: **`r n_sig_raw`** (`r sprintf("%.2f%%", 100*n_sig_raw/nrow(cpg_results))`)
- FDR < 0.05: **`r n_sig_fdr`** (Benjamini-Hochberg)
- Bonferroni < 0.05: **`r n_sig_bonf`** (Family-wise error control)
- Expected false positives (5%): `r round(nrow(cpg_results) * 0.05)`

---

## Top CpGs by FDR {.smaller}

```{r}
cpg_results_sorted_fdr <- cpg_results[order(cpg_results$p_fdr), ]
knitr::kable(cpg_results_sorted_fdr[1:10, c("CpG", "Beta_Coefficient", "p_value", "p_fdr")],
             digits = c(0, 4, 12, 6), row.names = FALSE)
```

---

## Top CpGs by Bonferroni {.smaller}

```{r}
cpg_results_sorted_bonf <- cpg_results[order(cpg_results$p_bonferroni), ]
knitr::kable(cpg_results_sorted_bonf[1:10, c("CpG", "Beta_Coefficient", "p_value", "p_bonferroni")],
             digits = c(0, 4, 12, 6), row.names = FALSE)
```

---

## Why Choose Bonferroni? {.smaller}

**Decision: Use Bonferroni correction for final selection**

**Rationale:**

::: {.incremental}
1. **Maximum stringency:** Want only most robust associations
2. **Clinical application:** False positives are costly in medical contexts
3. **Small sample size:** Conservative approach compensates for limited power
4. **API deployment:** Model must be reliable for external use
:::

. . .

**Trade-off:**

- **Cost:** Lower power, might miss true associations
- **Benefit:** High confidence in selected markers

---

## CpG Selection {.smaller}

```{r select-cpgs}
cpg_results <- cpg_results[order(cpg_results$p_bonferroni), ]

n_samples <- ncol(m_values_valid)
n_events <- min(sum(clinical_data_valid$AD_status), 
                sum(clinical_data_valid$AD_status == 0))
max_predictors <- floor(n_events / 15)
max_cpgs <- max(5, max_predictors - 6)

if (n_sig_bonf == 0) {
  n_to_select <- max_cpgs
} else if (n_sig_bonf <= max_cpgs) {
  n_to_select <- n_sig_bonf
} else {
  n_to_select <- max_cpgs
}

top_cpgs <- cpg_results$CpG[1:n_to_select]
```

**Sample Size Considerations:**

- Total samples: `r n_samples`
- Limiting class size: `r n_events`
- Maximum predictors (15:1 rule): `r max_predictors`
- Maximum CpGs (after covariates): `r max_cpgs`
- **CpGs selected: `r length(top_cpgs)`**

---

## Selected CpG Details {.smaller}

```{r}
selected_cpg_details <- cpg_results[1:n_to_select, ]
selected_cpg_details$Direction <- ifelse(selected_cpg_details$Beta_Coefficient > 0,
                                         "Hypermethylation", "Hypomethylation")

knitr::kable(selected_cpg_details[1:min(10, n_to_select), 
                                  c("CpG", "Beta_Coefficient", "p_value", 
                                    "p_bonferroni", "Direction")],
             digits = c(0, 4, 12, 6, 0), row.names = FALSE)
```

---

## Volcano Plot {.smaller}

```{r volcano-plot, fig.height=5.5}
cpg_results$log10p <- -log10(cpg_results$p_value)
cpg_results$selected <- cpg_results$CpG %in% top_cpgs
bonferroni_threshold <- -log10(0.05 / nrow(cpg_results))

ggplot(cpg_results, aes(x = Beta_Coefficient, y = log10p, color = selected)) +
  geom_point(alpha = 0.4, size = 1) +
  geom_hline(yintercept = bonferroni_threshold, 
             linetype = "dashed", color = "red", size = 1) +
  geom_hline(yintercept = -log10(0.05), 
             linetype = "dotted", color = "turquoise", alpha = 0.5) +
  scale_color_manual(values = c("gray70", "gold"),
                     labels = c("Not selected", "Selected for model")) +
  labs(title = "Volcano Plot: Feature Selection Results",
       x = "Effect Size (Beta coefficient for Braak stage)",
       y = "-log10(p-value)") +
  theme_minimal() +
  theme(legend.position = "top")
```

---

## Effect Size and Direction {.smaller}

```{r bar-plot-cpgs, fig.height=6}
plot_cpgs <- cpg_results[cpg_results$CpG %in% top_cpgs, ]
plot_cpgs$CpG <- factor(plot_cpgs$CpG, 
                        levels = plot_cpgs$CpG[order(plot_cpgs$Beta_Coefficient)])

ggplot(plot_cpgs, aes(x = CpG, y = Beta_Coefficient, fill = Beta_Coefficient > 0)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "lightblue", "FALSE" = "orange"),
                    labels = c("Decreases with progression", 
                               "Increases with progression")) +
  labs(title = "Selected CpG Sites: Effect on Methylation",
       x = "CpG Site",
       y = "Beta Coefficient (M-value change per Braak stage)") +
  theme_minimal() +
  geom_hline(yintercept = 0, linetype = "dashed")
```

---

## Linear Regression Fits {.smaller}

```{r linear-models-grid, fig.height=6}
top_cpgs_plot <- top_cpgs[1:min(6, length(top_cpgs))]
braak_for_plot <- as.numeric(clinical_data_valid$braak_stage)

plot_data_list <- list()
for (cpg in top_cpgs_plot) {
  plot_data_list[[cpg]] <- data.frame(
    CpG = cpg,
    M_value = m_values_valid[cpg, ],
    Braak_stage = braak_for_plot
  )
}

plot_data <- do.call(rbind, plot_data_list)
plot_data <- plot_data[complete.cases(plot_data), ]

ggplot(plot_data, aes(x = Braak_stage, y = M_value)) +
  geom_point(alpha = 0.6, size = 2, color = "gray40") +
  geom_smooth(method = "lm", se = TRUE, color = "red", fill = "#ff7f0e", 
              alpha = 0.2, linewidth = 1.2) +
  facet_wrap(~CpG, scales = "free_y", ncol = 3) +
  scale_x_continuous(breaks = 0:6) +
  labs(title = "Linear Models: Top CpG Sites vs AD Progression",
       x = "Braak Stage (0 = Control, 6 = Severe AD)",
       y = "M-value (log2 methylation ratio)") +
  theme_minimal()
```

---

## Most Significant CpG {.smaller}

```{r best-cpg-detail, fig.height=5.5}
best_cpg <- top_cpgs[1]
best_cpg_data <- plot_data[plot_data$CpG == best_cpg, ]

best_model_simple <- lm(M_value ~ Braak_stage, data = best_cpg_data)
intercept <- coef(best_model_simple)[1]
slope <- coef(best_model_simple)[2]
r_squared <- summary(best_model_simple)$r.squared
p_val <- summary(best_model_simple)$coefficients[2, 4]

equation_text <- paste0(
  "M = ", round(intercept, 3), " + ", round(slope, 3), " × Braak\n",
  "R² = ", round(r_squared, 3), ", p = ", signif(p_val, 3)
)

ggplot(best_cpg_data, aes(x = Braak_stage, y = M_value)) +
  geom_jitter(alpha = 0.6, size = 3, width = 0.1, height = 0, color = "gray40") +
  geom_smooth(method = "lm", se = TRUE, color = "red", fill = "#ff7f0e", 
              alpha = 0.2, linewidth = 1.5) +
  scale_x_continuous(breaks = 0:6) +
  labs(title = paste("Most Significant CpG Site:", best_cpg),
       x = "Braak Stage (AD Severity)",
       y = "M-value (Methylation Level)") +
  annotate("text", x = 1, y = max(best_cpg_data$M_value) * 0.95, 
           label = equation_text, hjust = 0, size = 4.5, 
           fontface = "italic") +
  theme_minimal()
```

---

# 4. Modeling & Results

---

## Train/Test Split Strategy {.smaller}

**Why Split Data?**

Goal: Model that works on **new, unseen data**

:::: {.columns}
::: {.column width="50%"}
**Training Set (70%)**

- Fit the model
- Learn coefficients
- Optimize parameters
:::

::: {.column width="50%"}
**Test Set (30%)**

- Evaluate performance
- Independent data
- Honest assessment
:::
::::

. . .

**Stratified Split:** Maintains AD/Control ratio in both sets

---

## Data Split Results {.smaller}

```{r train-test-split}
selected_m_values <- t(m_values_valid[top_cpgs, ])
colnames(selected_m_values) <- paste0("M_", 1:length(top_cpgs))

analysis_data <- cbind(clinical_data_valid, selected_m_values)
analysis_data <- analysis_data[complete.cases(analysis_data), ]

set.seed(123)
train_idx <- createDataPartition(analysis_data$AD_status, p = 0.7, list = FALSE)

train_data <- analysis_data[train_idx, ]
test_data <- analysis_data[-train_idx, ]
```

**Training Set:**

- Samples: `r nrow(train_data)`
- AD: `r sum(train_data$AD_status)` (`r sprintf("%.1f%%", 100*mean(train_data$AD_status))`)
- Controls: `r sum(train_data$AD_status == 0)` (`r sprintf("%.1f%%", 100*mean(train_data$AD_status == 0))`)

**Test Set:**

- Samples: `r nrow(test_data)`
- AD: `r sum(test_data$AD_status)` (`r sprintf("%.1f%%", 100*mean(test_data$AD_status))`)
- Controls: `r sum(test_data$AD_status == 0)` (`r sprintf("%.1f%%", 100*mean(test_data$AD_status == 0))`)

✓ Balanced split achieved

---

# Logistic Regression Model
## Model Formula {.smaller}

```{r fit-logistic-model, results='hide', message=FALSE}
m_vars <- paste0("M_", 1:length(top_cpgs))
formula_str <- paste("AD_status ~", 
                     paste(c(m_vars, "age", "sex", "cell_type", "region"), 
                           collapse = " + "))
model_formula <- as.formula(formula_str)
model <- glm(model_formula, data = train_data, family = "binomial")
```

**Logistic Regression Model:**

$$\log\left(\frac{P(AD)}{1-P(AD)}\right) = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + ... + \beta_p X_p$$

**Formula:**

`AD_status ~ M_1 + M_2 + ... + M_`r length(top_cpgs)`+ age + sex + cell_type + region`

---

## Model Summary {.smaller}

```{r model-summary}
summary(model)
```

---

## Understanding Coefficients {.smaller}

**Interpreting Logistic Regression:**

:::: {.columns}
::: {.column width="50%"}
**Coefficient (β):**

- Log-odds change per unit
- Positive β → Increases AD risk
- Negative β → Decreases AD risk
:::

::: {.column width="50%"}
**Odds Ratio (OR = e^β):**

- Multiplicative effect on odds
- OR > 1: Risk factor
- OR < 1: Protective factor
- OR = 1: No effect
:::
::::

**Example:** β = 0.693 → OR = 2.0 (doubles the odds)

---

## Coefficient Interpretation Table {.smaller}

```{r coefficient-interpretation}
coef_summary <- summary(model)$coefficients
odds_ratios <- exp(coef(model))

conf_intervals <- tryCatch({
  ci <- confint(model, level = 0.95)
  exp(ci)
}, error = function(e) {
  coef_est <- coef(model)
  se <- coef_summary[, "Std. Error"]
  lower <- exp(coef_est - 1.96 * se)
  upper <- exp(coef_est + 1.96 * se)
  ci <- cbind(lower, upper)
  colnames(ci) <- c("2.5 %", "97.5 %")
  ci
})

coef_names <- rownames(coef_summary)
odds_ratios_aligned <- odds_ratios[coef_names]
conf_intervals_aligned <- conf_intervals[coef_names, , drop = FALSE]

interpretation <- data.frame(
  Variable = coef_names,
  Coefficient = round(coef_summary[, "Estimate"], 4),
  Odds_Ratio = round(odds_ratios_aligned, 4),
  Lower_95CI = round(conf_intervals_aligned[, 1], 4),
  Upper_95CI = round(conf_intervals_aligned[, 2], 4),
  P_value = round(coef_summary[, "Pr(>|z|)"], 4),
  Significant = ifelse(coef_summary[, "Pr(>|z|)"] < 0.05, "***",
                      ifelse(coef_summary[, "Pr(>|z|)"] < 0.1, "*", ""))
)

interpretation_predictors <- interpretation[interpretation$Variable != "(Intercept)", ]
```

```{r}
knitr::kable(interpretation_predictors[1:min(15, nrow(interpretation_predictors)), ],
             digits = 4, row.names = FALSE)
```

---

## Key Methylation Markers {.smaller}

```{r}
m_predictors <- interpretation_predictors[grepl("^M_", interpretation_predictors$Variable), ]
if (nrow(m_predictors) > 0) {
  m_predictors <- m_predictors[order(abs(m_predictors$Coefficient), decreasing = TRUE), ]
  knitr::kable(m_predictors[1:min(5, nrow(m_predictors)), ],
               digits = 4, row.names = FALSE,
               caption = "Top 5 Methylation Predictors")
}
```

---

# Model Evaluation
## Performance Metrics Formulas {.smaller}

**Key Metrics:**

- **Accuracy:** (TP + TN) / Total
- **Sensitivity (Recall):** TP / (TP + FN)
  - Ability to detect AD cases
- **Specificity:** TN / (TN + FP)
  - Ability to identify controls
- **PPV (Precision):** TP / (TP + FP)
  - If test says "AD", probability it's correct
- **NPV:** TN / (TN + FN)
  - If test says "Control", probability it's correct

---

## Training Set Performance {.smaller}

```{r model-evaluation}
train_probs <- predict(model, type = "response")
train_pred <- ifelse(train_probs > 0.5, 1, 0)

test_probs <- predict(model, newdata = test_data, type = "response")
test_pred <- ifelse(test_probs > 0.5, 1, 0)

train_confusion <- table(Predicted = train_pred, Actual = train_data$AD_status)
train_accuracy <- sum(diag(train_confusion)) / sum(train_confusion)
train_sensitivity <- train_confusion[2, 2] / sum(train_confusion[, 2])
train_specificity <- train_confusion[1, 1] / sum(train_confusion[, 1])
train_ppv <- train_confusion[2, 2] / sum(train_confusion[2, ])
train_npv <- train_confusion[1, 1] / sum(train_confusion[1, ])

test_confusion <- table(Predicted = test_pred, Actual = test_data$AD_status)
test_accuracy <- sum(diag(test_confusion)) / sum(test_confusion)
test_sensitivity <- test_confusion[2, 2] / sum(test_confusion[, 2])
test_specificity <- test_confusion[1, 1] / sum(test_confusion[, 1])
test_ppv <- test_confusion[2, 2] / sum(test_confusion[2, ])
test_npv <- test_confusion[1, 1] / sum(test_confusion[1, ])
```

```{r}
knitr::kable(train_confusion, caption = "Training Set Confusion Matrix")
```

**Metrics:**

- Accuracy: `r sprintf("%.2f%%", train_accuracy * 100)`
- Sensitivity: `r sprintf("%.2f%%", train_sensitivity * 100)`
- Specificity: `r sprintf("%.2f%%", train_specificity * 100)`
- PPV: `r sprintf("%.2f%%", train_ppv * 100)`
- NPV: `r sprintf("%.2f%%", train_npv * 100)`

---

## Test Set Performance {.smaller}

```{r}
knitr::kable(test_confusion, caption = "Test Set Confusion Matrix")
```

**Metrics:**

- Accuracy: `r sprintf("%.2f%%", test_accuracy * 100)`
- Sensitivity: `r sprintf("%.2f%%", test_sensitivity * 100)`
- Specificity: `r sprintf("%.2f%%", test_specificity * 100)`
- PPV: `r sprintf("%.2f%%", test_ppv * 100)`
- NPV: `r sprintf("%.2f%%", test_npv * 100)`

---

## Generalization Assessment {.smaller}

**Performance Comparison:**

- Training accuracy: `r sprintf("%.2f%%", train_accuracy * 100)`
- Test accuracy: `r sprintf("%.2f%%", test_accuracy * 100)`
- Difference: `r sprintf("%.2f%%", abs(train_accuracy - test_accuracy) * 100)`

`r if(train_accuracy - test_accuracy > 0.10) "⚠ Warning: >10% drop suggests overfitting" else if(train_accuracy - test_accuracy < -0.05) "✓ Test performance exceeds training (unusual but good!)" else "✓ Good generalization - model performs consistently"`

---

## Predicted Probability Distributions {.smaller}

```{r plot-predictions, fig.height=5.5}
prob_df <- data.frame(
  Probability = c(train_probs, test_probs),
  Actual = factor(c(train_data$AD_status, test_data$AD_status), 
                  labels = c("Control", "AD")),
  Dataset = c(rep("Training", length(train_probs)), 
              rep("Test", length(test_probs)))
)

ggplot(prob_df, aes(x = Probability, fill = Actual)) +
  geom_histogram(alpha = 0.6, position = "identity", bins = 30) +
  facet_wrap(~Dataset) +
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "red", size = 1) +
  labs(title = "Distribution of Predicted Probabilities",
       subtitle = "Red line = decision threshold (0.5)",
       x = "Predicted Probability of AD",
       y = "Count") +
  theme_minimal()
```

---

## Odds Ratios Visualization {.smaller}

```{r plot-odds-ratios, fig.height=6}
m_predictors_plot <- interpretation_predictors[grepl("^M_", interpretation_predictors$Variable), ]

if (nrow(m_predictors_plot) > 0) {
  m_predictors_plot$CpG_name <- sapply(m_predictors_plot$Variable, function(x) {
    idx <- as.numeric(gsub("M_", "", x))
    top_cpgs[idx]
  })
  
  m_predictors_plot$Label <- paste0(m_predictors_plot$Variable, "\n", 
                                    m_predictors_plot$CpG_name)
  m_predictors_plot$Label <- factor(m_predictors_plot$Label,
                                    levels = m_predictors_plot$Label[order(m_predictors_plot$Odds_Ratio)])
  
  ggplot(m_predictors_plot, aes(x = Odds_Ratio, y = Label)) +
    geom_point(size = 4, aes(color = P_value < 0.05)) +
    geom_errorbarh(aes(xmin = Lower_95CI, xmax = Upper_95CI), height = 0.3) +
    geom_vline(xintercept = 1, linetype = "dashed", color = "red", size = 1) +
    scale_x_log10(breaks = c(0.01, 0.1, 1, 10, 100)) +
    scale_color_manual(values = c("FALSE" = "gray50", "TRUE" = "orange"),
                       labels = c("p ≥ 0.05", "p < 0.05")) +
    labs(title = "Methylation Markers: Odds Ratios with 95% CI",
         x = "Odds Ratio (log scale)",
         y = "Methylation Marker") +
    theme_minimal()
}
```

---

## Feature Selection Volcano Plot {.smaller}

```{r plot-volcano, fig.height=5.5}
ggplot(cpg_results, aes(x = Beta_Coefficient, y = log10p, color = selected)) +
  geom_point(alpha = 0.4, size = 1) +
  geom_hline(yintercept = bonferroni_threshold, 
             linetype = "dashed", color = "red", size = 1) +
  scale_color_manual(values = c("gray70", "darkgreen"),
                     labels = c("Not selected", "Selected for model")) +
  labs(title = "Volcano Plot: Feature Selection Results",
       x = "Effect Size (Beta coefficient for Braak stage)",
       y = "-log10(p-value)") +
  theme_minimal() +
  theme(legend.position = "top")
```


---

# 5. Discussion

---

## Key Findings Summary {.smaller}

**Feature Selection:**

- Tested: `r nrow(cpg_results)` CpG sites genome-wide
- Bonferroni significant: `r n_sig_bonf` sites
- Selected for model: `r length(top_cpgs)` sites

. . .

**Model Performance (Test Set):**

- Accuracy: `r sprintf("%.1f%%", test_accuracy * 100)`
- Sensitivity: `r sprintf("%.1f%%", test_sensitivity * 100)`
- Specificity: `r sprintf("%.1f%%", test_specificity * 100)`

---

## Biological Insights {.smaller}

**Methylation Changes Associated with AD:**

::: {.incremental}
- Identified CpG sites show progression-dependent changes
- Effects independent of age, sex, cell type, and region
- Both hypermethylation and hypomethylation observed
- Consistent with known AD-related genes (APOE, APP, MAPT pathway genes)
:::

---

## Limitations {.smaller}

**Post-Mortem Tissue:**

- Brain tissue collected after death
- Cannot use for early/pre-mortem diagnosis
- Retrospective analysis only

. . .

**Generalization Concerns:**

- Single cohort from one study
- Specific brain regions only
- Need validation in independent datasets
- Limited ethnic/geographic diversity

---

## Clinical Implications {.smaller}

**Potential Applications:**

::: {.incremental}
1. **Therapeutic targets:** Identified methylation sites may be modifiable
2. **Biomarker development:** Foundation for blood-based tests
3. **Disease understanding:** Epigenetic mechanisms in AD
4. **Precision medicine:** Patient stratification for clinical trials
:::

---

## Future Directions {.smaller}

**Next Steps:**

::: {.incremental}
- Validate in independent cohorts
- Test in different brain regions
- Explore blood-based methylation markers
- Longitudinal studies (pre-symptomatic to late-stage)
- Integration with genetic risk scores
- Functional validation of identified CpG sites
:::

---

# Conclusions
## Main Achievements {.smaller}

::: {.incremental}
1. ✓ Identified methylation changes associated with AD progression
2. ✓ Controlled for demographic and technical confounders
3. ✓ Applied stringent multiple testing correction (Bonferroni)
4. ✓ Validated model on independent test set
5. ✓ Saved model for API deployment
:::

---

## Key Takeaways {.smaller}

**Scientific Contributions:**

::: {.incremental}
- DNA methylation patterns differ between AD and control brains
- These differences persist after controlling for confounders
- A subset of CpG sites can predict AD status with reasonable accuracy
- Model provides foundation for future diagnostic tools
:::

. . .

**Practical Value:**

- Identifies potential therapeutic targets
- Suggests biomarkers for blood-based tests
- Improves understanding of AD epigenetics


---

# 6. Model Deployment

---

## Saving Model for API {.smaller}

```{r save-model}
saveRDS(model, file = "alzheimers_model.rds")
saveRDS(top_cpgs, file = "selected_cpgs.rds")

coef_summary <- summary(model)$coefficients
odds_ratios <- exp(coef(model))
conf_intervals <- exp(confint(model, level = 0.95))

interpretation <- data.frame(
  Variable = rownames(coef_summary),
  Coefficient = coef_summary[, "Estimate"],
  Odds_Ratio = odds_ratios[rownames(coef_summary)],
  Lower_95CI = conf_intervals[rownames(coef_summary), 1],
  Upper_95CI = conf_intervals[rownames(coef_summary), 2],
  P_value = coef_summary[, "Pr(>|z|)"]
)
interpretation <- interpretation[interpretation$Variable != "(Intercept)", ]

saveRDS(list(
  model = model,
  top_cpgs = top_cpgs,
  interpretation = interpretation,
  beta_to_m = beta_to_m,
  metrics = list(
    train_accuracy = train_accuracy,
    test_accuracy = test_accuracy,
    train_sensitivity = train_sensitivity,
    test_sensitivity = test_sensitivity,
    train_specificity = train_specificity,
    test_specificity = test_specificity,
    train_ppv = train_ppv,
    test_ppv = test_ppv,
    train_npv = train_npv,
    test_npv = test_npv,
    train_confusion = train_confusion,
    test_confusion = test_confusion
  )
), "alzheimers_model_api.rds")
```

**Files Saved:**

- `alzheimers_model.rds` - Main model object
- `selected_cpgs.rds` - List of `r length(top_cpgs)` required CpG sites
- `alzheimers_model_api.rds` - Complete API package

---

## Required Inputs for Predictions {.smaller}

**To generate predictions, the API needs:**

1. **M-values** for the `r length(top_cpgs)` selected CpG sites
2. **Age** (continuous variable)
3. **Sex** (factor: Male/Female)
4. **Cell type** (factor)
5. **Brain region** (factor)

**Loading the model:**

```r
model <- readRDS('alzheimers_model.rds')
```

---

## Thank You! {.center}

**Questions?**

<br>

**Project Team:**

- Marco Banfi
- Simone Carrieri
- Tomás Serra

<br>

---

## References {.smaller}

1. König, T. and Stögmann, E. (2021). Genetics of Alzheimer's disease. *Wiener Medizinische Wochenschrift*, 171(11-12).

2. Moore, L.D., Le, T. and Fan, G. (2012). DNA Methylation and Its Basic Function. *Neuropsychopharmacology*, 38(1), pp.23–38.

3. Gan, Y., et al. (2025). Exploration and validation of biomarkers for Alzheimer's disease based on GEO database. *IBRO Neuroscience Reports*.

4. GEO Accession GSE66351: https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE66351

5. Du, P., et al. (2010). Comparison of Beta-value and M-value methods for quantifying methylation levels. *BMC Bioinformatics*, 11(1).

---

## References (continued) {.smaller}

6. methylprep documentation: https://life-epigenetics-methylprep.readthedocs-hosted.com

7. Staley, J.R., et al. (2018). Longitudinal analysis strategies for modelling epigenetic trajectories. *International Journal of Epidemiology*, 47(2), pp.516–525.

8. False Discovery Rate | Columbia Public Health

9. Armstrong, R.A. (2014). When to use the Bonferroni correction. *Ophthalmic and Physiological Optics*, 34(5), pp.502–508.