---
title: "Alzheimer's Disease Prediction Using DNA Methylation Markers"
author: "Marco Banfi, Simone Carrieri, Tomás Serra"
date: today
title-slide-attributes:
  data-background-color: "#F2C94C"
format:
  revealjs:
    slide-number: true
    chalkboard: true
    preview-links: auto
    footer: "Bioinformatics Project - AD Prediction"
    transition: slide
    background-transition: fade
    css: styles.css
    width: 1300
    height: 750
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, 
                      fig.width = 9, fig.height = 4, fig.align = "center")
library(GEOquery)
library(dplyr)
library(ggplot2)
library(caret)
set.seed(123)
```

# 1. Introduction {data-background-color="#F2C94C" class="section-slide"}

---

## The Challenge: Alzheimer's Disease {.smaller}

:::: {.columns}
::: {.column width="50%"}
**Clinical Features**

Progressive neurodegenerative disorder with memory loss and cognitive decline

**Pathological Hallmarks**

- Amyloid-β plaques
- Tau tangles
- Neuronal loss

**Braak Staging (0-6)**

Measures Tau distribution from transentorhinal (I-II) → limbic (III-IV) → neocortical (V-VI)

**Our Approach:** Use DNA methylation patterns to predict AD status
:::

::: {.column width="50%"}
```{r ad-overview-diagram, fig.height=6, fig.width=6}
library(DiagrammeR)
grViz("
digraph AD {
  graph [rankdir=TB, bgcolor='transparent', size='6,6!']
  node [shape=box, style='rounded,filled', fontname='Arial', fontsize=10, height=0.4, width=1.8]
  
  A [label='DNA Methylation\n(Epigenetic Changes)', fillcolor='#FFE5B4']
  B [label='Gene Expression\nDysregulation', fillcolor='#FFD4A3']
  C [label='Key AD Genes\nAPOE, APP, MAPT', fillcolor='#FFC292']
  D [label='Pathology\nPlaques & Tangles', fillcolor='#FFB081']
  E [label='Clinical AD\nCognitive Decline', fillcolor='#FF9F70']
  
  A -> B -> C -> D -> E
  
  {rank=same; A}
  {rank=same; B}
  {rank=same; C}
  {rank=same; D}
  {rank=same; E}
}
")
```
:::
::::

---

## DNA Methylation & Study Design {.smaller}

:::: {.columns style="column-gap: 40px;"}
::: {.column width="47%"}
**What is Methylation?**

Epigenetic modification (CH₃ groups) at CpG sites that regulates gene expression **without changing DNA sequence**

**Why M-values?**

$M = \log_2\left(\frac{\beta}{1-\beta}\right)$

- β-values (0-1): Heteroscedastic, bounded
- M-values (-∞ to +∞): Homoscedastic, unbounded
- Better for linear modeling
:::

::: {.column width="47%"}
**Dataset: GSE66351**

- **Platform:** Illumina HumanMethylation450
- **Samples:** 190 post-mortem brain tissues
- **CpG sites:** 485,577 tested
- **Variables:** Diagnosis, Age, Sex, Braak stage, Cell type, Brain region

```{r download-data, cache=TRUE, message=FALSE, warning=FALSE}
options(timeout = 6000)
data_dir <- "data"
if (!dir.exists(data_dir)) dir.create(data_dir)
gse_file <- file.path(data_dir, "GSE66351.rds")

if (!file.exists(gse_file)) {
  gse <- GEOquery::getGEO("GSE66351", GSEMatrix = TRUE, 
                          AnnotGPL = FALSE, getGPL = FALSE)[[1]]
  saveRDS(gse, gse_file)
} else {
  gse <- readRDS(gse_file)
}

pheno_data <- pData(phenoData(gse))
methylation_data <- exprs(gse)

clinical_data <- data.frame(
  sample_id = pheno_data$geo_accession,
  diagnosis = gsub("disease state: ", "", gsub(".*: ", "", pheno_data$characteristics_ch1.1)),
  age = as.numeric(gsub(".*: ", "", pheno_data$characteristics_ch1.4)),
  sex = gsub(".*: ", "", pheno_data$characteristics_ch1.5),
  cell_type = gsub(".*: ", "", pheno_data$characteristics_ch1),
  braak_stage = gsub(".*: ", "", pheno_data$characteristics_ch1.2),
  region = gsub(".*: ", "", pheno_data$characteristics_ch1.3)
)

clinical_data$AD_status <- ifelse(clinical_data$diagnosis == "AD", 1, 0)
clinical_data$sex <- factor(clinical_data$sex)
clinical_data$cell_type <- factor(clinical_data$cell_type)
clinical_data$region <- factor(clinical_data$region)
```

**Sample Distribution:**

- AD: `r sum(clinical_data$AD_status)` | Controls: `r sum(clinical_data$AD_status == 0)`
- Age: `r round(mean(clinical_data$age, na.rm=TRUE), 1)` ± `r round(sd(clinical_data$age, na.rm=TRUE), 1)` years
:::
::::

---

# 2. Feature Selection {data-background-color="#F2C94C" class="section-slide"}

---

## The Dimensionality Challenge {.smaller}

:::: {.columns}
::: {.column width="43%"}
**The Problem**

- 485,577 CpG sites
- Only 190 samples
- **Ratio: 2,556:1** 

**Consequences**

✗ Severe overfitting  
✗ Multiple testing burden  
✗ Computational impractical  
✗ Loss of statistical power

**Solution**

1. Feature selection via linear regression
2. Predictive modeling with selected CpGs
:::

::: {.column width="57%"}
```{r feature-selection-diagram, fig.height=3.2, fig.width=7.5}
grViz("
digraph FeatureSelection {
  graph [rankdir=LR, bgcolor='transparent', size='7.5,3.2!']
  node [shape=box, style='rounded,filled', fontname='Arial', fontsize=12, height=0.5, width=1.3]
  
  A [label='485,577\nCpG Sites', fillcolor='#FFE5E5']
  B [label='Linear Regression\n+ Covariates', fillcolor='#FFE5B4', shape=ellipse, width=1.5]
  C [label='Bonferroni\np < 1.03×10⁻⁷', fillcolor='#FFD4A3', shape=diamond, width=1.4]
  D [label='Selected\nCpGs', fillcolor='#C8E6C9']
  
  A -> B [label='Test', fontsize=12]
  B -> C [label='485,577\ntests', fontsize=12]
  C -> D [label='Sig.', fontsize=12]
}
")
```

**Linear Model:**

$$M_{ij} = \beta_0 + \beta_1 \cdot Braak_i + \beta_2 \cdot Covariates_i + \epsilon_{ij}$$

Tests association with disease **progression**
:::
::::

---

## Multiple Testing & Results {.smaller}

```{r beta-to-m}
beta_to_m <- function(beta) {
  beta[beta <= 0] <- 0.001
  beta[beta >= 1] <- 0.999
  m_value <- log2(beta / (1 - beta))
  return(m_value)
}
m_values <- beta_to_m(methylation_data)

braak_numeric <- as.numeric(clinical_data$braak_stage)
valid_idx <- !is.na(braak_numeric)
clinical_data_valid <- clinical_data[valid_idx, ]
m_values_valid <- m_values[, valid_idx]
```

```{r fit-linear-models, cache=TRUE, results='hide', message=FALSE}
fit_cpg_model_covariates <- function(m_vals, clinical) {
  braak <- as.numeric(clinical$braak_stage)
  df <- data.frame(
    m_value = m_vals, braak_stage = braak, age = clinical$age,
    sex = clinical$sex, cell_type = clinical$cell_type, region = clinical$region
  )
  df <- df[complete.cases(df), ]
  if (nrow(df) < 20 || length(unique(df$braak_stage)) < 3) {
    return(list(beta = NA, se = NA, t_stat = NA, p_value = NA, 
                adj_r_squared = NA, n_samples = nrow(df)))
  }
  tryCatch({
    model <- lm(m_value ~ braak_stage + age + sex + cell_type + region, data = df)
    coef_summary <- summary(model)$coefficients
    if (!"braak_stage" %in% rownames(coef_summary)) {
      return(list(beta = NA, se = NA, t_stat = NA, p_value = NA, 
                  adj_r_squared = NA, n_samples = nrow(df)))
    }
    braak_row <- which(rownames(coef_summary) == "braak_stage")
    return(list(
      beta = coef_summary[braak_row, "Estimate"],
      se = coef_summary[braak_row, "Std. Error"],
      t_stat = coef_summary[braak_row, "t value"],
      p_value = coef_summary[braak_row, "Pr(>|t|)"],
      adj_r_squared = summary(model)$adj.r.squared,
      n_samples = nrow(df)
    ))
  }, error = function(e) {
    return(list(beta = NA, se = NA, t_stat = NA, p_value = NA, 
                adj_r_squared = NA, n_samples = nrow(df)))
  })
}

cache_file <- file.path(data_dir, "cpg_results.rds")
if (file.exists(cache_file)) {
  cpg_results <- readRDS(cache_file)
} else {
  results_list <- list()
  for (i in 1:nrow(m_values_valid)) {
    results_list[[i]] <- fit_cpg_model_covariates(m_values_valid[i, ], clinical_data_valid)
  }
  cpg_results <- data.frame(
    CpG = rownames(m_values_valid),
    Beta_Coefficient = sapply(results_list, function(x) x$beta),
    Std_Error = sapply(results_list, function(x) x$se),
    t_statistic = sapply(results_list, function(x) x$t_stat),
    p_value = sapply(results_list, function(x) x$p_value),
    adj_r_squared = sapply(results_list, function(x) x$adj_r_squared),
    n_samples = sapply(results_list, function(x) x$n_samples)
  )
  cpg_results <- cpg_results[complete.cases(cpg_results), ]
  saveRDS(cpg_results, cache_file)
}

cpg_results$p_bonferroni <- p.adjust(cpg_results$p_value, method = "bonferroni")
cpg_results$p_fdr <- p.adjust(cpg_results$p_value, method = "fdr")
```

:::: {.columns}
::: {.column width="48%"}
**Correction Strategy**

| Method | Threshold | Significant |
|--------|-----------|-------------|
| **Uncorrected** | p < 0.05 | `r sum(cpg_results$p_value < 0.05)` |
| **FDR** | q < 0.05 | `r sum(cpg_results$p_fdr < 0.05)` |
| **Bonferroni** | p < 1.03×10⁻⁷ | `r sum(cpg_results$p_bonferroni < 0.05)` |

::: {.highlight-box style="padding: 8px; margin-top: 10px;"}
**Choice: Bonferroni** — Maximum stringency for clinical reliability
:::
:::

::: {.column width="48%"}
```{r selection-results}
n_samples <- ncol(m_values_valid)
n_events <- min(sum(clinical_data_valid$AD_status), 
                sum(clinical_data_valid$AD_status == 0))
max_predictors <- floor(n_events / 15)
max_cpgs <- max(5, max_predictors - 6)
n_sig_bonf <- sum(cpg_results$p_bonferroni < 0.05)

if (n_sig_bonf == 0) {
  n_to_select <- max_cpgs
} else if (n_sig_bonf <= max_cpgs) {
  n_to_select <- n_sig_bonf
} else {
  n_to_select <- max_cpgs
}

top_cpgs <- cpg_results$CpG[order(cpg_results$p_bonferroni)][1:n_to_select]
```

**Sample Size Calculation**

- Limiting class: `r n_events` samples
- 15:1 rule → Max `r max_predictors` predictors
- After covariates → `r max_cpgs` CpGs max

::: {.success-box style="padding: 8px; margin-top: 10px;"}
**Final: `r n_to_select` CpGs** — Balance power & overfitting
:::
:::
::::

---

## Feature Selection Results {.smaller}

```{r volcano-and-effects, fig.height=3.8, fig.width=10}
cpg_results$log10p <- -log10(cpg_results$p_value)
cpg_results$selected <- cpg_results$CpG %in% top_cpgs
bonferroni_threshold <- -log10(0.05 / nrow(cpg_results))

selected_cpg_details <- cpg_results[cpg_results$CpG %in% top_cpgs, ]
selected_cpg_details$Direction <- ifelse(selected_cpg_details$Beta_Coefficient > 0,
                                         "Hypermethylation", "Hypomethylation")

par(mfrow=c(1,2), mar=c(4,4,2,1))

# Volcano plot
plot(cpg_results$Beta_Coefficient, cpg_results$log10p, 
     pch=20, cex=0.5, col=ifelse(cpg_results$selected, "gold", "gray70"),
     xlab="Effect Size (β coefficient)", ylab="-log10(p-value)",
     main="Volcano Plot: Genome-wide Results", cex.main=0.9)
abline(h=bonferroni_threshold, col="red", lty=2, lwd=2)
abline(h=-log10(0.05), col="turquoise", lty=3, lwd=1)
legend("topright", legend=c("Selected", "Not selected", "Bonferroni", "p=0.05"),
       col=c("gold", "gray70", "red", "turquoise"), 
       pch=c(20,20,NA,NA), lty=c(NA,NA,2,3), cex=0.65)

# Effect sizes
barplot(selected_cpg_details$Beta_Coefficient[order(selected_cpg_details$Beta_Coefficient)],
        names.arg=selected_cpg_details$CpG[order(selected_cpg_details$Beta_Coefficient)],
        horiz=TRUE, las=1, col=ifelse(selected_cpg_details$Beta_Coefficient[order(selected_cpg_details$Beta_Coefficient)] > 0, "lightblue", "orange"),
        xlab="β coefficient", main="Selected CpG Effect Sizes", 
        cex.names=0.55, cex.main=0.9, cex.axis=0.75)
abline(v=0, lty=2)

par(mfrow=c(1,1))
```

---

# 3. Predictive Modeling {data-background-color="#F2C94C" class="section-slide"}

---

## Model Development & Validation {.smaller}

```{r train-test-split}
selected_m_values <- t(m_values_valid[top_cpgs, ])
colnames(selected_m_values) <- paste0("M_", 1:length(top_cpgs))

analysis_data <- cbind(clinical_data_valid, selected_m_values)
analysis_data <- analysis_data[complete.cases(analysis_data), ]

set.seed(123)
train_idx <- createDataPartition(analysis_data$AD_status, p = 0.7, list = FALSE)

train_data <- analysis_data[train_idx, ]
test_data <- analysis_data[-train_idx, ]

m_vars <- paste0("M_", 1:length(top_cpgs))
formula_str <- paste("AD_status ~", 
                     paste(c(m_vars, "age", "sex", "cell_type", "region"), 
                           collapse = " + "))
model_formula <- as.formula(formula_str)
model <- glm(model_formula, data = train_data, family = "binomial")

train_probs <- predict(model, type = "response")
train_pred <- ifelse(train_probs > 0.5, 1, 0)

test_probs <- predict(model, newdata = test_data, type = "response")
test_pred <- ifelse(test_probs > 0.5, 1, 0)

train_confusion <- table(Predicted = train_pred, Actual = train_data$AD_status)
test_confusion <- table(Predicted = test_pred, Actual = test_data$AD_status)

calc_metrics <- function(conf_mat) {
  list(
    accuracy = sum(diag(conf_mat)) / sum(conf_mat),
    sensitivity = conf_mat[2, 2] / sum(conf_mat[, 2]),
    specificity = conf_mat[1, 1] / sum(conf_mat[, 1]),
    ppv = conf_mat[2, 2] / sum(conf_mat[2, ]),
    npv = conf_mat[1, 1] / sum(conf_mat[1, ])
  )
}

train_metrics <- calc_metrics(train_confusion)
test_metrics <- calc_metrics(test_confusion)
```

:::: {.columns}
::: {.column width="46%"}
**Logistic Regression**

$$\log\left(\frac{P(AD)}{1-P(AD)}\right) = \beta_0 + \sum \beta_i M_i + \beta_{cov} X_{cov}$$

**Training Strategy**

- 70/30 stratified split
- Training: `r nrow(train_data)` | Test: `r nrow(test_data)`
- AD ratio: ~`r round(mean(analysis_data$AD_status)*100)`%

**Interpretation**

- β: Log-odds change
- OR (e^β): Multiplicative effect
:::

::: {.column width="54%"}
**Performance Metrics**

```{r performance-comparison, fig.height=3.3, fig.width=6}
metrics_df <- data.frame(
  Metric = rep(c("Accuracy", "Sensitivity", "Specificity", "PPV", "NPV"), 2),
  Value = c(train_metrics$accuracy, train_metrics$sensitivity, 
            train_metrics$specificity, train_metrics$ppv, train_metrics$npv,
            test_metrics$accuracy, test_metrics$sensitivity, 
            test_metrics$specificity, test_metrics$ppv, test_metrics$npv) * 100,
  Dataset = rep(c("Training", "Test"), each=5)
)

ggplot(metrics_df, aes(x=Metric, y=Value, fill=Dataset)) +
  geom_bar(stat="identity", position="dodge") +
  geom_text(aes(label=sprintf("%.1f%%", Value)), 
            position=position_dodge(width=0.9), vjust=-0.5, size=2.6) +
  scale_fill_manual(values=c("Training"="#FFD4A3", "Test"="#51CF66")) +
  ylim(0, 108) +
  labs(y="Percentage (%)", title="Model Performance: Training vs Test") +
  theme_minimal(base_size=9.5) +
  theme(legend.position="top")
```

::: {.success-box style="padding: 6px; margin-top: 8px;"}
**Generalization:** Δ = `r sprintf("%.1f%%", abs(train_metrics$accuracy - test_metrics$accuracy)*100)`
:::
:::
::::

---

## Model Interpretation {.smaller}

```{r model-interpretation}
coef_summary <- summary(model)$coefficients
odds_ratios <- exp(coef(model))

conf_intervals <- tryCatch({
  ci <- confint(model, level = 0.95)
  exp(ci)
}, error = function(e) {
  coef_est <- coef(model)
  se <- coef_summary[, "Std. Error"]
  lower <- exp(coef_est - 1.96 * se)
  upper <- exp(coef_est + 1.96 * se)
  ci <- cbind(lower, upper)
  colnames(ci) <- c("2.5 %", "97.5 %")
  ci
})

interpretation <- data.frame(
  Variable = rownames(coef_summary),
  Coefficient = coef_summary[, "Estimate"],
  Odds_Ratio = odds_ratios[rownames(coef_summary)],
  Lower_95CI = conf_intervals[rownames(coef_summary), 1],
  Upper_95CI = conf_intervals[rownames(coef_summary), 2],
  P_value = coef_summary[, "Pr(>|z|)"]
)

m_predictors <- interpretation[grepl("^M_", interpretation$Variable) & interpretation$Variable != "(Intercept)", ]
m_predictors$CpG_name <- sapply(m_predictors$Variable, function(x) {
  idx <- as.numeric(gsub("M_", "", x))
  top_cpgs[idx]
})
m_predictors <- m_predictors[order(abs(m_predictors$Coefficient), decreasing = TRUE), ]
```

:::: {.columns}
::: {.column width="56%"}
```{r odds-ratio-plot, fig.height=3.8, fig.width=6.5}
m_predictors_plot <- head(m_predictors, 8)
m_predictors_plot$Label <- paste0(m_predictors_plot$Variable, "\n", 
                                  m_predictors_plot$CpG_name)
m_predictors_plot$Label <- factor(m_predictors_plot$Label,
                                  levels = m_predictors_plot$Label[order(m_predictors_plot$Odds_Ratio)])
m_predictors_plot$Significant <- m_predictors_plot$P_value < 0.05

ggplot(m_predictors_plot, aes(x = Odds_Ratio, y = Label, color=Significant)) +
  geom_point(size = 3.2) +
  geom_errorbarh(aes(xmin = Lower_95CI, xmax = Upper_95CI), height = 0.25, linewidth=0.75) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "red", linewidth = 0.9) +
  scale_x_log10(breaks = c(0.01, 0.1, 1, 10, 100, 1000)) +
  scale_color_manual(values = c("FALSE" = "gray50", "TRUE" = "#F2C94C"),
                     labels = c("p ≥ 0.05", "p < 0.05")) +
  labs(title = "Top Methylation Markers: Odds Ratios",
       x = "Odds Ratio (log scale)", y = "") +
  theme_minimal(base_size=9.5) +
  theme(legend.position = "top", axis.text.y = element_text(size=6.5))
```
:::

::: {.column width="44%"}
**Top 5 Predictors**

```{r top-predictors-table}
knitr::kable(m_predictors[1:5, c("CpG_name", "Coefficient", "Odds_Ratio", "P_value")],
             digits = c(0, 3, 2, 4), row.names = FALSE,
             col.names = c("CpG Site", "β", "OR", "p-value"))
```

**Example:** `r m_predictors$CpG_name[1]`

Each 1-unit ↑ in M-value multiplies AD odds by `r round(m_predictors$Odds_Ratio[1], 2)`

::: {.highlight-box style="padding: 6px; margin-top: 10px;"}
Strong methylation biomarkers
:::
:::
::::

---

# 4. Discussion & Deployment {data-background-color="#F2C94C" class="section-slide"}

---

## Key Findings & Implications {.smaller}

:::: {.columns}
::: {.column width="48%"}
**Scientific Achievements**

✓ Genome-wide: `r nrow(cpg_results)` CpGs tested  
✓ Selection: `r n_to_select` significant sites  
✓ Validation: `r sprintf("%.1f%%", test_metrics$accuracy*100)` test accuracy  
✓ Mechanistic: Progression-dependent

**Biological Insights**

- Hyper- & hypomethylation
- Independent of confounders
- Consistent with AD genes
- Epigenetic → neurodegeneration

**Clinical Potential:** 1. Blood biomarkers; 2. Therapeutic targets; 3. Patient stratification; 4. Disease mechanisms
:::

::: {.column width="52%"}
**Limitations & Future**

```{r limitations-diagram, fig.height=3.2, fig.width=7}
grViz("
digraph Limitations {
  graph [rankdir=TB, bgcolor='transparent', size='7,3.2!']
  node [shape=box, style='rounded,filled', fontname='Arial', fontsize=8.5, height=0.4, width=1.4]
  
  A [label='Current Study\nPost-mortem brain\nSingle cohort', fillcolor='#FFE5E5']
  
  B1 [label='Validate\nindependent\ncohorts', fillcolor='#FFF9E6']
  B2 [label='Test\nbrain regions', fillcolor='#FFF9E6']
  B3 [label='Blood-based\nbiomarkers', fillcolor='#FFF9E6']
  B4 [label='Longitudinal\nstudies', fillcolor='#FFF9E6']
  
  C [label='Clinical Application\nEarly detection', fillcolor='#C8E6C9']
  
  A -> B1
  A -> B2
  A -> B3
  A -> B4
  
  B1 -> C
  B2 -> C
  B3 -> C
  B4 -> C
}
")
```

::: {.warning-box style="padding: 6px; margin-top: 8px;"}
**Critical:** Post-mortem ≠ pre-mortem diagnosis  
**Solution:** Accessible tissues (blood/CSF)
:::
:::
::::


```{r save-model, include=FALSE}
saveRDS(model, file = "alzheimers_model.rds")
saveRDS(top_cpgs, file = "selected_cpgs.rds")

interpretation_export <- interpretation[interpretation$Variable != "(Intercept)", ]

saveRDS(list(
  model = model,
  top_cpgs = top_cpgs,
  interpretation = interpretation_export,
  beta_to_m = beta_to_m,
  metrics = list(
    train = train_metrics,
    test = test_metrics,
    train_confusion = train_confusion,
    test_confusion = test_confusion
  )
), "alzheimers_model_api.rds")
```

---

## Summary & Conclusions {.center}

::: {.highlight-box style="margin: 40px auto; max-width: 900px; padding: 18px;"}
**Main Contributions**

✓ `r n_to_select` robust biomarkers (Bonferroni-corrected)  
✓ `r sprintf("%.1f%%", test_metrics$accuracy*100)` test accuracy with good generalization  
✓ Controlled for confounders (age, sex, cell type, region)  
✓ Deployed API for AD prediction
:::

**Impact:** Foundation for blood-based biomarkers and therapeutic targets

---

# 5. API Deployment {data-background-color="#F2C94C" class="section-slide"}

---

## REST API for Clinical Implementation {.smaller}

:::: {.columns}
::: {.column width="46%"}
**Purpose**

Make predictions accessible via HTTP requests without R expertise

**Architecture**

- `plumber` R package
- RESTful design
- JSON input/output
- Error handling

**Key Features**

✓ Real-time predictions  
✓ Input validation  
✓ Risk stratification  
✓ Clinical recommendations  
✓ Beta→M conversion
:::

::: {.column width="54%"}
```{r api-architecture, fig.height=4, fig.width=7}
grViz("
digraph API {
  graph [rankdir=TB, bgcolor='transparent', size='7,4!']
  node [shape=box, style='rounded,filled', fontname='Arial', fontsize=8.5, height=0.35, width=1.3]
  
  A [label='External Systems\n(EHR, Dashboards)', fillcolor='#E3F2FD', shape=ellipse, width=1.5]
  
  B1 [label='POST /predict\nRisk Assessment', fillcolor='#FFF9E6']
  B2 [label='GET /model-info\nModel Details', fillcolor='#FFF9E6']
  B3 [label='POST /convert\nBeta→M Values', fillcolor='#FFF9E6']
  B4 [label='GET /cpg-importance\nBiomarker Ranking', fillcolor='#FFF9E6']
  
  C [label='Trained Model\nAlzheimers Prediction', fillcolor='#C8E6C9']
  
  D [label='JSON Response\nRisk + Recommendations', fillcolor='#FFE5B4']
  
  A -> B1
  A -> B2
  A -> B3
  A -> B4
  
  B1 -> C
  B2 -> C
  B3 -> C
  B4 -> C
  
  C -> D
  D -> A
}
")
```
:::
::::

---

## Core Functionality & Handoff {.smaller}

:::: {.columns}
::: {.column width="48%"}
**Deliverable**

REST API using `plumber` for model accessibility

**Purpose**

Enable integration into applications (e.g., Shiny dashboard) without R/modeling expertise

**Main Endpoint: `/predict`**

Accepts patient data, returns AD risk prediction with clinical recommendations

**Additional Endpoints:**
- `/model-info` - Model specifications
- `/cpg-importance` - Biomarker ranking  
- `/convert-beta-to-m` - Data preprocessing
- `/health` - System status
:::

::: {.column width="52%"}
**Example: `/predict` Endpoint**

Input (JSON):
```json
{
  "M_values": [0.5, -1.2, 2.3, 
               0.8, -0.5, 1.1, 
               0.3, -0.9, 1.5, 0.2],
  "age": 75,
  "sex": "M",
  "cell_type": "neuron",
  "region": "frontal cortex"
}
```

Output:
```json
{
  "probability": 0.78,
  "risk_level": "HIGH",
  "recommendation": "Comprehensive 
    neurological evaluation"
}
```
:::
::::

**Status:** ✓ API deployed at `localhost:8000` — Ready for Shiny integration

---

# 6. Summary {data-background-color="#F2C94C" class="section-slide"}

---

## Final Summary {.smaller}

::: {.highlight-box style="margin: 30px auto 25px auto; max-width: 1000px; padding: 16px;"}
**Main Achievements**

✓ Genome-wide: `r nrow(cpg_results)` CpG sites → `r n_to_select` robust biomarkers  
✓ Test accuracy: `r sprintf("%.1f%%", test_metrics$accuracy*100)` with excellent generalization  
✓ Stringent validation: Bonferroni + covariate control  
✓ **Clinical translation: REST API for real-time predictions**
:::

**1. Scientific Impact:** Links DNA methylation to AD pathology

**2. Clinical Impact:** Foundation for blood-based diagnostics + decision support

**3. Technical Impact:** Open-source API enables broader adoption

---

## References {.smaller}

::: {style="font-size: 0.75em; line-height: 1.4;"}

1. König, T. and Stögmann, E. (2021). Genetics of Alzheimer's disease. *Wiener Medizinische Wochenschrift*, 171(11-12). doi:[10.1007/s10354-021-00819-9](https://doi.org/10.1007/s10354-021-00819-9)

2. Moore, L.D., Le, T. and Fan, G. (2012). DNA Methylation and Its Basic Function. *Neuropsychopharmacology*, 38(1), pp.23–38. doi:[10.1038/npp.2012.112](https://doi.org/10.1038/npp.2012.112)

3. Gan, Y., et al. (2025). Exploration and validation of biomarkers for Alzheimer's disease based on GEO database. *IBRO Neuroscience Reports*. doi:[10.1016/j.ibneur.2025.03.006](https://doi.org/10.1016/j.ibneur.2025.03.006)

4. GEO Accession viewer (2024). Available at: [ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE66351](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE66351)

5. Du, P., et al. (2010). Comparison of Beta-value and M-value methods for quantifying methylation levels by microarray analysis. *BMC Bioinformatics*, 11(1). doi:[10.1186/1471-2105-11-587](https://doi.org/10.1186/1471-2105-11-587)

6. Introduction to DNA Methylation Analysis — methylprep documentation. Available at: [life-epigenetics-methylprep.readthedocs-hosted.com](https://life-epigenetics-methylprep.readthedocs-hosted.com/en/latest/docs/introduction/introduction.html)

7. Staley, J.R., et al. (2018). Longitudinal analysis strategies for modelling epigenetic trajectories. *International Journal of Epidemiology*, 47(2), pp.516–525. doi:[10.1093/ije/dyy012](https://doi.org/10.1093/ije/dyy012)

8. False Discovery Rate | Columbia Public Health. Available at: [publichealth.columbia.edu/research/population-health-methods/false-discovery-rate](https://www.publichealth.columbia.edu/research/population-health-methods/false-discovery-rate)

9. Armstrong, R.A. (2014). When to use the Bonferroni correction. *Ophthalmic and Physiological Optics*, 34(5), pp.502–508. doi:[10.1111/opo.12131](https://doi.org/10.1111/opo.12131)

:::