---
title: "BioInfProjectPhase2"
author: "Marco Banfi, Simone Carrieri, Tomás Serra"
format:
  pdf:
    pdf-engine: xelatex
    toc: true
    number-sections: true
    colorlinks: true
    linkcolor: blue
    urlcolor: blue
    citecolor: blue
    geometry:
      - margin=2.5cm
    fontsize: 11pt
    documentclass: article
    keep-tex: true
editor: visual
---

\newpage

# **Introduction**

We first performed a short issue-management step to identify and prioritize the main open issues in the exchanged project. Then Phase 2 extends the Phase 1 predictive service (Plumber API + trained logistic regression model) by adding a Shiny user interface that consumes the API strictly as a black box, and by providing a fully reproducible deployment using two interconnected Docker containers (API + Shiny) orchestrated with Docker Compose. The Phase 1 model and inference logic remain unchanged. The UI collects patient inputs, performs basic client-side validation, sends a JSON request to `/predict`, and presents the returned death probability clearly (including a simple risk bucket for presentation only), while the API is responsible for loading the model artifact and producing predictions.

***Deliverable Summary***

We provide:

-   a Shiny user interface (UI) that collects patient inputs, validates them, and display prediction results.

-   a Dockerized two-container deployment (API + UI) orchestrated via Docker Compose

- A minimal verification protocol (health check + example prediction) for users

# **System Overview**

## **Components and responsabilities**

**API (Plumber container)**

- Loads the model artifact at startup.

- Exposes:

  - `GET /health` (readiness check)
  
  - `POST /predict` (inference)

**UI (Shiny container)**

- Collects patient input variables.

- Performs basic client-side validation.

- Sends a JSON request to the API and displays the returned probability.

- Does not contain or load the model artifact.

**Architecture diagram**

```{mermaid}
%%| fig-align: center
%%| fig-width: 5.4
%%| fig-height: 5.0
%%| fig-cap: "Phase 2 architecture: user accesses Shiny UI and can also reach the API directly; the UI calls the API over the Docker network; the API loads the model at startup."
%%{init: {
  "flowchart": {
    "nodeSpacing": 26,
    "rankSpacing": 34,
    "curve": "step"
  },
  "themeVariables": {
    "fontSize": "10.5px",
    "padding": 6
  }
}}%%
flowchart TB
  U["User"]

  subgraph UI["UI container"]
    S["Shiny UI (app.R)"]
  end

  subgraph API["API container"]
    A["Plumber API (plumber.R)"]
    M["Model (.rds)"]
  end

  U -->|3838| S
  U -->|8000| A
  S -->|api:8000| A
  A -->|loads| M



```

## Networking semantics (Host vs Docker network)

From the host machine (your Windows terminal/browser):

- API is reachable at `http://localhost:8000`
- Shiny is reachable at `http://localhost:3838`

From inside Docker (container-to-container):

- The Shiny container reaches the API at `http://api:8000`

Here, `api` is the service name defined in `docker-compose.yml` and is resolvable via Docker DNS.

## **Project Structure**

``` text
.
+-- api/
|   +-- Dockerfile
|   +-- plumber.R
|   +-- model/
|       +-- model_logistic_DEATH_EVENT.rds
+-- shiny/
|   +-- app.R
|   +-- Dockerfile
+-- docker-compose.yml
+-- README.md
```


# API Specification (Plumber)

## Endpoints

| Endpoint   | Method | Purpose                  | Success Response                   |
| ---------- | -----: | ------------------------ | ---------------------------------- |
| `/health`  |    GET | Service readiness check  | JSON: `{ "status": "ok" }`         |
| `/predict` |   POST | Return death probability | JSON: `{ "probability": <float> }` |


## Input contract for `/predict`

The request body is JSON with the following fields:

| Field               | Type    | Expected range (server validation) |
| ------------------- | ------- | ---------------------------------- |
| `age`               | numeric | 0–120                              |
| `ejection_fraction` | numeric | 0–100                              |
| `serum_creatinine`  | numeric | >0–20                              |
| `serum_sodium`      | numeric | 90–200                             |

On invalid inputs, the API returns HTTP 400 with a JSON error message, e.g. `{ "error": "..." }`.


## Swagger / API docs

Plumber interactive documentation (when enabled by Plumber) is available at:

- `http://localhost:8000/__docs__/`

# Shiny UI Behavior

The Shiny UI is designed to be a black-box client:

- It never loads `.rds` files and does not perform model inference locally.

- It only communicates with the API via HTTP and depends on the API contract.

Core UI actions:

- Read numeric inputs aligned to the API contract.

- Validate inputs locally (fast feedback to users).

- Send JSON to `/predict`.

- Display:

  - predicted probability (as percent formatting)

  - a simple presentation-only risk bucket (not a clinical threshold)

## API base URL configuration

The UI reads `API_BASE_URL`:

- Inside Docker Compose: `API_BASE_URL=http://api:8000`

- Local UI without Docker networking: `API_BASE_URL=http://127.0.0.1:8000`

# Containerization and Deployment

## API container (`api/Dockerfile`)

- Base image: `rocker/r-ver:4.3.1`

- Installs system dependencies and R packages (`plumber`, `jsonlite`).

- Copies `plumber.R` and `model/` into `/app/`.

- Exposes port `8000`.

- Starts Plumber on `0.0.0.0:8000`.

## Shiny container (`shiny/Dockerfile`)

- Base image: `rocker/shiny:4.3.1`

- Installs R packages for HTTP + JSON handling.

- Copies `app.R` into the Shiny app directory.

- Exposes port `3838`.

## Orchestration (`docker-compose.yml`)

Docker Compose:

- Builds two images (API and Shiny) from separate contexts (`./api` and `./shiny`).

- Creates a shared Docker network for inter-service communication.

- Maps host ports:

  - `8000:8000` for the API

  - `3838:3838` for Shiny

# How to Run and Verify

In this project, the **primary execution and verification** were performed in **GitHub Codespaces** using Docker Compose and port forwarding. **Local execution is supported**, but in our case it was not feasible due to limited disk/storage on the local machine.

### Primary workflow — GitHub Codespaces (recommended)

**Step 1 - Start services**

From the repository root (where `docker-compose.yml` is located):

```bash
docker compose up -d --build
```

**Step 2 - Check containers status**

```bash
docker compose ps
```

```{r fig-docker-ps, echo=FALSE, results="asis", fig.cap="Docker Compose running state (`docker compose ps -a`).", out.width="90%", fig.align="center"}
f <- "figures/docker_compose_ps.png"
if (file.exists(f)) {
  knitr::include_graphics(f)
} else {
  cat("Error: Add a screenshot of docker compose ps -a (saved as figures/docker_compose_ps.png).\n")
}
```

**Step 3 - Forward ports**

In the Codespaces **Ports** panel, forward:

- `8000` (API)
- `3838` (Shiny)

```{r fig-codespaces-ports, echo=FALSE, results="asis", fig.cap="GitHub Codespaces port forwarding for the API (8000) and Shiny UI (3838).", out.width="90%", fig.align="center"}

f <- "figures/codespaces_ports_forwarding.png"
if (file.exists(f)) {
  knitr::include_graphics(f)
} else {
  cat("Error: Add a screenshot of Codespaces port forwarding (saved as figures/codespaces_ports_forwarding.png).\n")
}
```

### Step 4 — Verify the API

Using the forwarded URL for port `8000`:

- Open `/health` for readiness

- Open `/__docs__/` for Swagger UI

```{r fig-swagger, echo=FALSE, results="asis", fig.cap="Swagger UI generated by the Plumber API (/__docs__/) showing /health and /predict.", out.width="90%", fig.align="center"}

f <- "figures/swagger_docs_screenshot.png"
if (file.exists(f)) {
  knitr::include_graphics(f)
} else {
  cat("Error: Add a Swagger screenshot (saved as figures/swagger_docs_screenshot.png).\n")
}
```

(Example: `<FORWARDED_8000_URL>/health`)

### Step 5 — Verify the UI

Using the forwarded URL for port `3838`:

- Open the Shiny application (root)

- If your setup serves the app under a subpath, use `/app`

(Example: `<FORWARDED_3838_URL>/app`)


## API-level tests (Codespaces or local terminal)

If you run the commands inside the Codespaces terminal, you can still test using `localhost` because the terminal runs in the same environment where containers are running.

Health check:

```bash
curl http://localhost:8000/health
```

Prediction example:

```bash
curl -X POST http://localhost:8000/predict -H "Content-Type: application/json" -d '{"age":60,"ejection_fraction":40,"serum_creatinine":1.2,"serum_sodium":135}'
```

Expected example response:

```json
{"probability":[0.1857]}
```

## UI verification (browser)

Open the Shiny UI via the forwarded port `3838` URL and run one prediction to verify:

- inputs are validated locally,

- the API status is reachable,

- a probability is returned and displayed.

***Shiny UI — Predict Tab: Patient inputs and prediction output (probability + risk bucket)***

```{r fig-shiny-predict, echo=FALSE, results="asis", fig.cap="Shiny UI — Predict Tab: Patient inputs and prediction output (probability + risk bucket).", out.width="90%", fig.align="center"}

f <- "figures/shiny_ui_screenshot1.png"
if (file.exists(f)) {
  knitr::include_graphics(f)
} else {
  cat("Error: Add a Shiny UI screenshot (saved as figures/shiny_ui_screenshot.png).\n")
}
```

***Shiny UI — Model & Dataset Tab: Model predictors and API contract (/predict, /health)***

```{r fig-shiny-model-dataset, echo=FALSE, results="asis", fig.cap="Shiny UI — Model & Dataset Tab: Model predictors and API contract (/predict, /health).", out.width="90%", fig.align="center"}

f <- "figures/shiny_ui_screenshot2.png"
if (file.exists(f)) {
  knitr::include_graphics(f)
} else {
  cat("Error: Add a Shiny UI screenshot (saved as figures/shiny_ui_screenshot.png).\n")
}
```

\newpage

***Shiny UI — Deployment Tab: Docker Compose run/test commands (curl + stop)***

```{r fig-shiny-deployment, echo=FALSE, results="asis", fig.cap="Shiny UI — Deployment Tab: Docker Compose run/test commands (curl + stop).", out.width="90%", fig.align="center"}

f <- "figures/shiny_ui_screenshot3.png"
if (file.exists(f)) {
  knitr::include_graphics(f)
} else {
  cat("Error: Add a Shiny UI screenshot (saved as figures/shiny_ui_screenshot.png).\n")
}
```



## Local execution (optional)

Local execution follows the same commands, but requires sufficient disk space to pull large base images and build containers.

Run:

``` bash
docker compose up -d --build
```

Stop:

``` bash
docker compose down
```

Open in browser (host machine):

- API: `http://localhost:8000/health` and `http://localhost:8000/__docs__/`

- Shiny: `http://localhost:3838/app`


## Challenges and Solutions

### Limited local storage for Docker images

**Issue:** Local execution was not feasible due to limited disk/storage on the local machine (large base images and build layers).  
**Resolution:** We used GitHub Codespaces as the primary execution environment. Codespaces allowed us to build and run the two-container deployment reliably, and to expose both services via port forwarding (`8000` for the API, `3838` for Shiny).

### UI shows “API status unreachable”

Common causes and fixes:

- **API container not running:** verify with `docker compose ps`.

- **Wrong base URL:** inside Docker use `http://api:8000`, on host use `http://127.0.0.1:8000`.

- **Port forwarding not enabled (Codespaces):** forward ports `8000` and `3838` from the Ports panel.

- **Port conflicts (local):** ensure `8000`/`3838` are not used by other applications.

