---
title: "BioInfProjectPhase2"
author: "Marco Banfi, Simone Carrieri, Tomás Serra"
format:
  pdf:
    pdf-engine: xelatex
    toc: true
    number-sections: true
    colorlinks: true
    linkcolor: blue
    urlcolor: blue
    citecolor: blue
    geometry:
      - margin=2.5cm
    fontsize: 11pt
    documentclass: article
    keep-tex: true
editor: visual
---

\newpage

# **Introduction**

Option 1) We first carried out a short issue management step to identify the main open issues in the exchanged project and prioritize the fixes. We then extend the Phase 1 predictive service (Plumber API + trained logistic regression model) by adding a Shiny user interface and a fully reproducible deployment based on two interconnected Docker containers (API + Shiny). We keep the Phase 1 model and inference logic unchanged: in Phase 2, we treat the API as a black box and focus on usability, clean integration, and portability across environments. The UI collects patient inputs, performs basic validation, sends a JSON request to /predict, and presents the returned death probability clearly (including a simple risk bucket for presentation), while the API remains responsible for loading the unchanged model from model/model_logistic_DEATH_EVENT.rds and producing predictions.

Option 2) Phase 2 extends the Phase 1 predictive service (Plumber API + trained logistic regression model) by adding a Shiny user interface that consumes the API strictly as a black box, and by providing a reproducible deployment with two interconnected Docker containers (API + Shiny). After the issue management phase, no changes are introduced to the trained model or the inference logic; the focus is on usability, integration, and deployment reproducibility.

***Deliverable Summary*** 

In this deliverable we provide:

- a Shiny web UI that collects patient inputs, calls the Plumber API, and displays the predicted death probability (with a simple risk bucket for presentation purposes);

- a documented API contract for `/health` and `/predict` (including input validation rules and JSON payload/response format);

- a reproducible deployment via two Docker containers (API + Shiny) orchestrated with `docker compose`;

- a practical execution and verification workflow (GitHub Codespaces with port forwarding).

# **Architecture**

## **System Overview**

The system is composed of:

-   **API service (plumber, Phase 1)**\
    Runs `plumber.R`, loads `model/model_logistic_DEATH_EVENT.rds`, and exposes REST endpoints.
-   **UI service (Shiny, Phase 2)**\
    Runs `shiny/app.R`, collects inputs, calls the API, and renders the returned prediction.
-   **Orchestration (Docker Compose)**\
    `docker-compose.yml` builds, starts, and networks the two services.

## **Architecture Diagram**

```{mermaid}
%%| fig-align: center
%%| fig-width: 5.4
%%| fig-height: 5.0
%%| fig-cap: "Phase 2 architecture: user accesses Shiny UI and can also reach the API directly; the UI calls the API over the Docker network; the API loads the model at startup."
%%{init: {
  "flowchart": {
    "nodeSpacing": 26,
    "rankSpacing": 34,
    "curve": "step"
  },
  "themeVariables": {
    "fontSize": "10.5px",
    "padding": 6
  }
}}%%
flowchart TB
  U["User"]

  subgraph UI["UI container"]
    S["Shiny UI (app.R)"]
  end

  subgraph API["API container"]
    A["Plumber API (plumber.R)"]
    M["Model (.rds)"]
  end

  U -->|3838| S
  U -->|8000| A
  S -->|api:8000| A
  A -->|loads| M



```

## **Networking Semantics: Host vs Container**

-   From the host machine (your browser/terminal), http://localhost:8000 works because Docker publishes port 8000 to the host.

-   From inside the Shiny container, localhost refers to the Shiny container itself. Therefore, the Shiny app must call the API using the Docker Compose service name:

-   Inside Docker: API_BASE_URL = http://api:8000

-   Running Shiny locally: API_BASE_URL = http://127.0.0.1:8000

## **Project Structure**

```text
.
+-- api/
|   +-- Dockerfile
+-- shiny/
|   +-- app.R
|   +-- Dockerfile
+-- docker-compose.yml
+-- plumber.R
+-- model/
|   +-- model_logistic_DEATH_EVENT.rds
+-- README.md
```

# **How to Run (Docker)**

From the repository root:

``` bash
docker compose up -d --build
```

Then open:

-   API: `http://localhost:8000`

-   Swagger: `http://localhost:8000/__docs__/`

-   Shiny UI: `http://localhost:3838/app`

```{r fig-docker-ps, echo=FALSE, results="asis", fig.cap="Docker Compose running state (`docker compose ps -a`).", out.width="90%", fig.align="center"}
f <- "figures/docker_compose_ps.png"
if (file.exists(f)) {
  knitr::include_graphics(f)
} else {
  cat("Error: Add a screenshot of docker compose ps -a (saved as figures/docker_compose_ps.png).\n")
}
```

## **Running from GitHub Codespaces (No Local Docker Required)**

The full system can be executed in GitHub Codespaces and accessed through forwarded ports. (save locally storage space!)

1)  Create a Codespace

-   Open the repository on GitHub

-   Click Code → Codespaces

-   Click Create codespace on main

-   Wait for the VS Code environment to load

2)  Start the System in Codespaces

In the Codespaces terminal:

``` bash
docker compose up -d --build
```

3)  Forward Ports and Access Services

In the Codespaces Ports panel:

Ensure ports 8000 (API) and 3838 (Shiny) are forwarded

Open the forwarded URLs and append:

API health check: `/health`

Swagger: `/__docs__/`

Shiny UI: `/app`

```{r fig-codespaces-ports, echo=FALSE, results="asis", fig.cap="GitHub Codespaces port forwarding for the API (8000) and Shiny UI (3838).", out.width="90%", fig.align="center"}

f <- "figures/codespaces_ports_forwarding.png"
if (file.exists(f)) {
  knitr::include_graphics(f)
} else {
  cat("Error: Add a screenshot of Codespaces port forwarding (saved as figures/codespaces_ports_forwarding.png).\n")
}
```

4)  Stop Services and Release Resources

``` bash
docker compose down
```

## **API Tests (Docker / Codespaces)**

### **Health Check**

``` bash
curl http://localhost:8000/health
```

*Expected response:*

``` json
{"status":"ok"}
```

### **Prediction**

``` bash
curl -i -X POST http://localhost:8000/predict \
  -H "Content-Type: application/json" \
  -d '{"age":60,"ejection_fraction":40,"serum_creatinine":1.2,"serum_sodium":135}'
```

*Expected response fields:*

`probability` (numeric in `[0, 1]`)

# **Swagger API Documentation**

Swagger is available at:

- `http://localhost:8000/__docs__/` (local Docker)

- forwarded port URL + `/__docs__/` (Codespaces)

```{r fig-swagger, echo=FALSE, results="asis", fig.cap="Swagger UI generated by the Plumber API (/__docs__/) showing /health and /predict.", out.width="90%", fig.align="center"}

f <- "figures/swagger_docs_screenshot.png"
if (file.exists(f)) {
  knitr::include_graphics(f)
} else {
  cat("Error: Add a Swagger screenshot (saved as figures/swagger_docs_screenshot.png).\n")
}
```

# **Phase 2 UI Behavior (Black-Box Consumption)**

The Shiny UI:

-   collects the required clinical inputs:

    -   `age`, `ejection_fraction`, `serum_creatinine`, `serum_sodium`

-   performs basic client-side range validation

-   checks API availability via GET /health and displays a status indicator

-   sends inputs as JSON to POST /predict

-   renders the returned probability and a presentation-friendly percentage

-   optionally displays the raw JSON response and allows downloading the latest prediction as a JSON file

The UI does not execute the model locally and contains no inference logic beyond validation and presentation.

\newpage

***Shiny UI — Predict Tab: Patient inputs and prediction output (probability + risk bucket)***

```{r fig-shiny-predict, echo=FALSE, results="asis", fig.cap="Shiny UI — Predict Tab: Patient inputs and prediction output (probability + risk bucket).", out.width="90%", fig.align="center"}

f <- "figures/shiny_ui_screenshot1.png"
if (file.exists(f)) {
  knitr::include_graphics(f)
} else {
  cat("Error: Add a Shiny UI screenshot (saved as figures/shiny_ui_screenshot.png).\n")
}
```

***Shiny UI — Model & Dataset Tab: Model predictors and API contract (/predict, /health)***

```{r fig-shiny-model-dataset, echo=FALSE, results="asis", fig.cap="Shiny UI — Model & Dataset Tab: Model predictors and API contract (/predict, /health).", out.width="90%", fig.align="center"}

f <- "figures/shiny_ui_screenshot2.png"
if (file.exists(f)) {
  knitr::include_graphics(f)
} else {
  cat("Error: Add a Shiny UI screenshot (saved as figures/shiny_ui_screenshot.png).\n")
}
```

\newpage

***Shiny UI — Deployment Tab: Docker Compose run/test commands (curl + stop)***

```{r fig-shiny-deployment, echo=FALSE, results="asis", fig.cap="Shiny UI — Deployment Tab: Docker Compose run/test commands (curl + stop).", out.width="90%", fig.align="center"}

f <- "figures/shiny_ui_screenshot3.png"
if (file.exists(f)) {
  knitr::include_graphics(f)
} else {
  cat("Error: Add a Shiny UI screenshot (saved as figures/shiny_ui_screenshot.png).\n")
}
```



# **Some challenges & Solutions**

**Docker build fails (R packages missing)**

If logs show errors such as:

-   `there is no package called 'plumber'`

-   `there is no package called 'httr'`

the Docker build likely missed system-level Linux dependencies required to compile/install R packages. Ensure Dockerfiles install OS dependencies before `install.packages()`, e.g.:

-   `build-essential`, `pkg-config`

-   `libcurl4-openssl-dev`, `libssl-dev`, `libxml2-dev`

-   `libsodium-dev` (for package `sodium`)

-   `zlib1g-dev` (for compilation dependencies)

Rebuild without cache:

``` bash
docker compose down
docker compose build --no-cache api shiny
docker compose up -d
```

Check logs:

``` bash
docker compose logs --tail=120 api
docker compose logs --tail=120 shiny
```

**UI shows API unreachable**

Verify both containers are running:

``` bash
docker compose ps -a
```

Inside Docker, the UI must call the API using `http://api:8000` (not `localhost`).

**Hostname api cannot be resolved**

This happens when running the Shiny app outside Docker. Use Docker Compose, or set:

-   `API_BASE_URL=http://127.0.0.1:8000`

**Port already in use**

If ports 8000 or 3838 are occupied on the host, stop the conflicting process or change mappings in `docker-compose.yml` (e.g., `"8001:8000"`).

**`curl` shows malformed input / bad hostname**

This typically comes from incorrect line breaks. Use a single-line `curl` command or ensure `\` is the final character on each continued line.

**Shiny fails to start after editing app.R**

A frequent cause is an R parsing error (e.g., unbalanced quotes). Inspect logs:

``` bash
docker compose logs -f shiny
```



